---
title: "Simulating Complex Traits"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---


Load packages and define folders
```{r}
##install.packages("tidyverse")
library(tidyverse)
data.dir = "/home/haky/IntroStatGen/data/hapmap/"
bin.dir = "/home/haky/IntroStatGen/bin/"
out.dir ="/home/haky/IntroStatGen/out/"
gist.dir = "/home/haky/IntroStatGen/repo-IntroStatGen/gists/"

plinkname = paste0(bin.dir,"plink")
header0 = "founders-common" ## EUR-common, hapmap-common, hapmap2-common
bedheader = paste0(data.dir,header0)

seednumber = 1
outheader = paste0(out.dir,header0,"-",seednumber)

##sample prop % of SNPs to make them causal
propC = 0.10

## run from scratch if header0 changes
from_scratch = FALSE

```

calculate GRM
```{r}

## calculate GRM
if(from_scratch)
{gctafile = paste0(bin.dir,"gcta64")
comm = paste0(gctafile, " --bfile ", bedheader," --make-grm-bin --out ", outheader)
print(comm)
system(comm)
}

## calculate grm left using ch1 to ch8 
if(from_scratch)
{gctafile = paste0(bin.dir,"gcta64")
comm = paste0(gctafile, " --bfile ", bedheader,"-left --make-grm-bin --out ", outheader,"-left")
print(comm)
system(comm)
}

## calculate grm right using ch9 to ch22 
if(from_scratch)
{gctafile = paste0(bin.dir,"gcta64")
comm = paste0(gctafile, " --bfile ", bedheader,"-right --make-grm-bin --out ", outheader,"-right")
print(comm)
system(comm)
}
```

Read marker information from plinkfile.bim
```{r}
##read hapmap.bim

bimdata = read_tsv(paste0(bedheader, ".bim"),col_names = FALSE)
names(bimdata) = c("chr","rsid","dis","pos","A1","A2")
head(bimdata)
Mt = nrow(bimdata)
MC = round(propC * Mt) ## number of causal SNPs

```


Randomly choose propC% SNPs as causal
```{r, eval=FALSE}

set.seed(seednumber) ## we want to get the same SNPs each time we run this
indC = sample(1:Mt, MC) ## select Mc SNPs to be simulated as causal

```

Simulate betas as normally distributed for the propC% of SNPs selected above
```{r simulate betas, eval=FALSE}

set.seed(seednumber + 1)
betaC = rnorm(MC,mean=0, sd=1/sqrt(MC))
hist(betaC)

## generate weights file, Y = sum(weight * X) + error; we call beta or weight interchangeably
weightdata = bimdata[indC,] %>% select(rsid,A1) 
weightdata$betas = betaC
weightfile = paste0(out.dir, "weights-seed-",seednumber,".txt")
write_tsv(weightdata,path=weightfile)

```

Calculate the genetic component using plink --score
```{r}
## run plink -score
comm = paste0(bin.dir,"plink --bfile ", bedheader," --score ", weightfile, " header sum --out ", out.dir,"polyscore-seed",seednumber)
print(comm)
system( comm )

## read sum(beta * X) from polyscore-seed1.profile
PSdata = read_table2(paste0(out.dir,"polyscore-seed",seednumber,".profile"))
head(PSdata)
hist(PSdata$SCORESUM)
```

Calculate var of epsilon as a function of variance of beta and heritability
```{r}

h2 = 0.60
sigepsi2 = (1-h2)/h2 * var(PSdata$SCORESUM)
print(h2)
print(sigepsi2,2)
print(var(PSdata$SCORESUM)/(var(PSdata$SCORESUM)+sigepsi2))

## homework problem: show analytically that variance of epsilon defined above leads to the heritability h2. 
## i.e. that h2 = var(sum(beta X)) / var(Y); knowing that var(Y) = var(sum(beta X)) + var(epsilon)

```

Calculate phenotype and write file to be used with plink
```{r}

phenofile = paste0(outheader,".txt")
nsamp = nrow(PSdata)
set.seed(seednumber + 2); ## use seednumber + 2 to get reproducible simulations
epsilon = rnorm(nsamp,mean=0,sd=sqrt(sigepsi2) )
PSdata = PSdata %>% mutate(PHENO = SCORESUM + epsilon)
write_tsv(PSdata %>% select(FID,IID,PHENO,SCORESUM),path=paste0(phenofile))

## homework problem: show empirically that the heritability of the phenotype is = h2
## hint var(SCORESUM) / var(PHENO) 

```

Run GWAS with the phenotype
```{r }

comm = paste0(plinkname, " --bfile ", bedheader, " --assoc --pheno ",phenofile, " --pheno-name PHENO ", " --out ", outheader)
print(comm)
system(comm)
```


Plots histogram of p values
```{r}

qassoc = read_table(paste0(outheader, ".qassoc"))
dim(qassoc)
head(qassoc)
hist(qassoc$P)
source(paste0(gist.dir, "qqunif.r")) ## load qqunif function from the gist folder
qqunif.plot(qassoc$P,main = paste0("sim with h2 = ",round(h2,2)) )

```

```{r, echo=FALSE}
##install.packages("qqman")
library(qqman)
ind=sample(1:nrow(assocPC), 1e5) ## select subset to plot. We wouldn't do this in a real analysis since we could be missing significant hits by subsampling.
manhattan(qassoc[ind,])

```

compare true betas with estimates
Notice 90% of betas are 0
plot estimated betas vs true betas
```{r}

## inner join qassoc and weights
tempo = qassoc %>% rename(betahat = BETA)%>% left_join(weightdata %>% select(rsid,betas),by=c("SNP"="rsid")) 
tempo = tempo %>% mutate(beta = ifelse(is.na(betas),0,betas)) ## if not in weights file, beta=0
## plot
tempo %>% sample_frac(0.10) %>% ggplot(aes(betas, betahat)) + geom_point() + geom_abline()

## small region plot beta vs position

## LD

```




calculate heritability

```{r}
grmcutoff = 1 ## this is no cutoff. typically one would use --grm-cutoff 0.025

## calculate h2 using all chromosomes
comm = paste0(gctafile, " --reml ", " --grm ", outheader, " --pheno ", phenofile ,"  --grm-cutoff ",grmcutoff, " --out ", outheader, "-gcta-all")
print(comm)
system(comm)

## calculate heritability using chr 1 to chr 8)
comm = paste0(gctafile, " --reml ", " --grm ", outheader, "-left --pheno ", phenofile , " --grm-cutoff ",grmcutoff," --out ", outheader, "-gcta-left")
print(comm)
system(comm)

## calculate heritability using chr 9 to chr 22)
comm = paste0(gctafile, " --reml ", " --grm ", outheader, "-right --pheno ", phenofile , " --grm-cutoff ",grmcutoff," --out ", outheader, "-gcta-right")
print(comm)
system(comm)

```



```{r}

## generate PCs with gcta (above we generated pcs with plink but need to get rid of header and I only want 5 pcs. Once the GRM is calculated, PCs are cheap to generate: 1.3 seconds)


comm = paste0(gctafile, " --pca 5 ", " --grm ", outheader , " --out ", outheader, "-gcta-covar")
print(comm)
system(comm)


```


Read GRM

```{r}

titulo = "all pop GRM first row"
grmmat = OmicKriging::read_GRMBin(paste0("/home/haky/IntroStatGen/out/simpheno-all-1.grm"))
tempo = data.frame(IID=names(grmmat[1,]),cor=grmmat[1,])
tempo = tempo %>% inner_join(popinfo,by=c("IID"="IID"))
hist(tempo$cor,main=titulo)
tempo %>% filter(cor<1) %>% ggplot(aes(IID,cor,col=population,label=population)) + geom_point() + geom_label() + ggtitle(titulo)

titulo = "EUR GRM first row"
grmmat = OmicKriging::read_GRMBin(paste0("/home/haky/IntroStatGen/out/simpheno-EUR-1.grm"))
tempo = data.frame(IID=names(grmmat[1,]),cor=grmmat[1,])
tempo = tempo %>% inner_join(popinfo,by=c("IID"="IID"))
hist(tempo$cor,main=titulo)
tempo %>% filter(cor<.99) %>% ggplot(aes(IID,cor,col=population,label=population)) + geom_point() + geom_label()+ ggtitle(titulo)




```




use Yang formula to adjust if necessary

